Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2019-08-05 analysis/paper/Rmd/hotspot-one-chromosome.Rmd
fixed chunk name by removing underscores
removed calls to save figures

```{r, scatter-{{hot_chr}}, fig.cap="Pleiotropy test statistic vs. mediation LOD difference for Chr {{hot_chr}} hotspot."}
p <- pleio_mixed_list %>%
  bind_rows() %>%
  rename(pleiotropy_lod = lod) %>%
  inner_join(med_tib, by = c("local_gene_id", "nonlocal_gene_id")) %>%
  dplyr::select(local_symbol, nonlocal_symbol, pleiotropy_lod, lod_diff, lod_diff_proportion, lod_without_med, lod_with_med, filename, nonlocal_gene_id, local_gene_id) %>%
  ggplot() + geom_point(mapping = aes(x = pleiotropy_lod, y = lod_diff, colour = local_symbol == "{{keller_mediator}}", nonlocal = nonlocal_symbol, local = local_symbol), size = 0.05) + theme(legend.position = "none") + xlab("Pleiotropy LOD") + ylab("Mediation LOD difference")
fnsvg <- paste0("Chr", {{hot_chr}}, "_scatter.svg")
fnpdf <- paste0("Chr", {{hot_chr}}, "_scatter.pdf")
ggsave(plot = p, filename = here("analysis", "figures", fnsvg))
ggsave(plot = p, filename = here("analysis", "figures", fnpdf))
plotly::ggplotly(p)
```

```{r, scatter_by_local-{{hot_chr}}, fig.height = ceiling(nlocal / 2), fig.cap="Pleiotropy test statistic vs. mediation LOD difference by local expression trait for Chr {{hot_chr}} hotspot."}
foo <- pleio_mixed_list %>%
  bind_rows() %>%
  rename(pleiotropy_lod = lod) %>%
  inner_join(med_tib, by = c("local_gene_id", "nonlocal_gene_id")) %>%
  dplyr::select(local_symbol, nonlocal_symbol, pleiotropy_lod, lod_diff, lod_diff_proportion, lod_without_med, lod_with_med, filename, nonlocal_gene_id, local_gene_id)
p <- foo %>%
  ggplot() +
  geom_point(mapping = aes(x = pleiotropy_lod, y = lod_diff, colour = nonlocal_symbol), size = 0.1) +
  facet_wrap(. ~ local_symbol, nrow = ceiling(nlocal / 2), ncol = 2) +
  broman::karl_theme(strip.placement = "inside") +
  theme(legend.position = "none") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
) +
  xlab("Pleiotropy LOD") +
  ylab("Mediation LOD difference") +
  geom_text(aes(x = max(foo$pleiotropy_lod), y = 1, label = local_symbol), inherit.aes=FALSE)

fnsvg <- paste0("Chr", {{hot_chr}}, "_scatter-panel.svg")
fnpdf <- paste0("Chr", {{hot_chr}}, "_scatter-panel.pdf")
ggsave(p, filename = here("analysis", "figures", fnsvg), height = ceiling(nlocal / 2))
ggsave(p, filename = here("analysis", "figures", fnpdf), height = ceiling(nlocal / 2))
plotly::ggplotly(p)
```

```{r, eval = FALSE}
# put in multiple figures, if there are many local traits
n_figs <- 1 + nlocal %/% 18 # 18 is the number of panels per page in pdf
n_last <- nlocal %% 18 # number of panels on last figure
max_pleio_lod <- max(foo$pleiotropy_lod)
max_med_lod_difference <- max(foo$lod_diff)
min_med_lod_difference <- min(foo$lod_diff)

for (i in 1:n_figs){
  n_panels <- 18 * (i < n_figs) + n_last * (i == n_figs)
  foo %>%
    filter(local_symbol %in% local_annot2$symbol[((i - 1) * 18 + 1): (18 * i)]) %>%
    ggplot() +
      geom_point(mapping = aes(x = pleiotropy_lod, y = lod_diff, colour = nonlocal_symbol), size = 0.1) +
      facet_wrap(. ~ local_symbol, nrow = ceiling(n_panels / 2), ncol = 2) +
      broman::karl_theme() +
      theme(legend.position = "none") +
      theme(
        strip.background = element_blank(),
        strip.text.x = element_blank()
    ) +
      xlab("Pleiotropy LOD") + xlim(c(0, max_pleio_lod)) + ylim(c(min_med_lod_difference, max_med_lod_difference)) +
      ylab("Mediation LOD difference") +
      geom_text(size = 2, aes(x = max_pleio_lod / 2, y = 2 * max_med_lod_difference / 3, label = local_symbol), inherit.aes=FALSE)
  fnsvg <- paste0("Chr", {{hot_chr}}, "_scatter-panel", i, ".svg")
  fnpdf <- paste0("Chr", {{hot_chr}}, "_scatter-panel", i, ".pdf")
  ggsave(filename = here("analysis", "figures", fnsvg), height = ceiling(n_panels / 2))
  ggsave(filename = here("analysis", "figures", fnpdf), height = ceiling(n_panels / 2))
}
```

@@

```{r, ordering-{{hot_chr}}}
# order the nonlocal traits first
## indicator for being a nonlocal trait
indic_nonlocal <- rownames(pm) %in% nonlocal_annot$symbol
# define ord_nonlocal
ord_nonlocal <- pm[indic_nonlocal, indic_nonlocal] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[indic_nonlocal, indic_nonlocal])) %>%
  order.dendrogram()
rn_nonlocal <- rownames(pm[indic_nonlocal, indic_nonlocal][ord_nonlocal, ord_nonlocal])
# define ord_local
ord_local<- pm[!indic_nonlocal, !indic_nonlocal] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[!indic_nonlocal, !indic_nonlocal])) %>%
  order.dendrogram()
rn_local <- rownames(pm[!indic_nonlocal, !indic_nonlocal][ord_local, ord_local])
rn <- c(rn_nonlocal, rn_local)
```

@@

```{r, n_med-{{hot_chr}}}
n_med_tr <- 4 # arbitrary threshold for number of mediated traits per local
# only those local traits with n_med_tr or more mediated nonlocal traits 
# appear as annotation columns in the heatmap
n_med_per_local <- med_tib %>%
  group_by(local_symbol) %>%
  filter(lod_diff > 1.5) %>%
  count() %>%
  filter(n >= n_med_tr)
n_med_per_local %>%
  arrange(desc(n)) %>%
  knitr::kable()

scols_pre <- scols_extra %>%
  dplyr::select(n_med_per_local$local_symbol)
# get ordering for columns of scols_pre
col_order <- scols_pre %>%
  t() %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(t(scols_pre))) %>%
  order.dendrogram()
scols <- scols_pre[ , col_order] %>%
  mutate(local = scols1)
fnhtml <- here("analysis", "figures", paste0("Chr", {{hot_chr}}, "hm.html"))
```

```{r, heat-{{hot_chr}}, fig.cap="Heatmap for pleiotropy test statsitics for Chr {{hot_chr}}."}
ord <- match(rn, rownames(pm))

(p <- pm[ord, ord] %>%
  apply(FUN = trunc2, MARGIN = 2, threshold = 5) %>%
  heatmaply(Rowv = FALSE,
            Colv = FALSE,
            cexRow = 0.2,
            cexCol = 0.2,
            symm = TRUE,
            row_side_colors = scols[ord, ],
            col_side_colors = tibble(local = scols1[ord]),
            limits = c(0, 5)
            ) %>%
  hide_guides()
)
fnpdf <- here("analysis", "figures", paste0("Chr", {{hot_chr}}, "hm.pdf"))
htmlwidgets::saveWidget(widget = p, 
                        file = fnhtml
                        )
```

```{r, webshot-{{hot_chr}}, eval = FALSE}
webshot::webshot(url = fnhtml,
                 file = fnpdf,
                 cliprect = "viewport"
                 )
```

@@

p <- tibble(nonlocal_symbol = rownames(pm[indic_nonlocal, indic_nonlocal])[ord_nonlocal], index = 1:length(nonlocal_symbol)) %>%
  inner_join(bar, by = "nonlocal_symbol") %>% 
  left_join(nonlocal_annot, by = c("nonlocal_symbol" = "symbol")) %>%
  ggplot() + geom_point(aes(y = pos, x = index, color = lod_diff > 1.5, nonlocal = nonlocal_symbol, lod_diff = lod_diff))

>>>>

```{r, scatter-{{hot_chr}}, fig.cap="Pleiotropy test statistic vs. mediation LOD difference for Chr {{hot_chr}} hotspot."}
p <- pleio_mixed_list %>%
  bind_rows() %>%
  rename(pleiotropy_lod = lod) %>%
  inner_join(med_tib, by = c("local_gene_id", "nonlocal_gene_id")) %>%
  dplyr::select(local_symbol, nonlocal_symbol, pleiotropy_lod, lod_diff, lod_diff_proportion, lod_without_med, lod_with_med, filename, nonlocal_gene_id, local_gene_id) %>%
  ggplot() + geom_point(mapping = aes(x = pleiotropy_lod, y = lod_diff, colour = local_symbol == "{{keller_mediator}}", nonlocal = nonlocal_symbol, local = local_symbol), size = 0.05) + theme(legend.position = "none") + xlab("Pleiotropy LOD") + ylab("Mediation LOD difference")
plotly::ggplotly(p)
```

```{r, scatter-by-local-{{hot_chr}}, fig.height = ceiling(nlocal / 2), fig.cap="Pleiotropy test statistic vs. mediation LOD difference by local expression trait for Chr {{hot_chr}} hotspot."}
foo <- pleio_mixed_list %>%
  bind_rows() %>%
  rename(pleiotropy_lod = lod) %>%
  inner_join(med_tib, by = c("local_gene_id", "nonlocal_gene_id")) %>%
  dplyr::select(local_symbol, nonlocal_symbol, pleiotropy_lod, lod_diff, lod_diff_proportion, lod_without_med, lod_with_med, filename, nonlocal_gene_id, local_gene_id)
p <- foo %>%
  ggplot() +
  geom_point(mapping = aes(x = pleiotropy_lod, y = lod_diff, colour = nonlocal_symbol), size = 0.1) +
  facet_wrap(. ~ local_symbol, nrow = ceiling(nlocal / 2), ncol = 2) +
  broman::karl_theme(strip.placement = "inside") +
  theme(legend.position = "none") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
) +
  xlab("Pleiotropy LOD") +
  ylab("Mediation LOD difference") +
  geom_text(aes(x = max(foo$pleiotropy_lod), y = 1, label = local_symbol), inherit.aes=FALSE)
plotly::ggplotly(p)
```

@@

```{r, ordering-{{hot_chr}}}
# order the nonlocal traits first
## indicator for being a nonlocal trait
indic_nonlocal <- rownames(pm) %in% nonlocal_annot$symbol
# define ord_nonlocal
ord_nonlocal <- pm[indic_nonlocal, ] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[indic_nonlocal, ])) %>%
  order.dendrogram()
rn_nonlocal <- rownames(pm[indic_nonlocal, ][ord_nonlocal, ])
# define ord_local
ord_local<- pm[!indic_nonlocal, ] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[!indic_nonlocal, ])) %>%
  order.dendrogram()
rn_local <- rownames(pm[!indic_nonlocal, ][ord_local, ])
rn <- c(rn_nonlocal, rn_local)
```

@@

```{r, n-med-{{hot_chr}}}
n_med_tr <- 10 # arbitrary threshold for number of mediated traits per local
# only those local traits with n_med_tr or more mediated nonlocal traits 
# appear as annotation columns in the heatmap
n_med_per_local <- med_tib %>%
  group_by(local_symbol) %>%
  filter(lod_diff > 1.5) %>%
  count() %>%
  filter(n >= n_med_tr)
n_med_per_local %>%
  arrange(desc(n)) %>%
  knitr::kable(caption = "Number of mediated traits per local trait for Chr {{hot_chr}} hotspot.")

scols_pre <- scols_extra %>%
  dplyr::select(n_med_per_local$local_symbol)
# get ordering for columns of scols_pre
col_order <- scols_pre %>%
  t() %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(t(scols_pre))) %>%
  order.dendrogram()
scols <- scols_pre[ , col_order] %>%
  mutate(local = scols1)
```

```{r, heat-{{hot_chr}}, fig.cap="Heatmap for pleiotropy test statistics for Chr {{hot_chr}}."}
ord <- match(rn, rownames(pm))

(p <- pm[ord, ord] %>% # order both cols and rows by 'ord'
  apply(FUN = trunc2, MARGIN = 2, threshold = 5) %>%
  heatmaply(Rowv = FALSE,
            Colv = FALSE,
            cexRow = 0.2,
            cexCol = 0.2,
            symm = TRUE,
            row_side_colors = scols[ord, ],
            col_side_colors = tibble(local = scols1[ord]),
            limits = c(0, 5)
            ) %>%
  hide_guides()
)
```

@@

p <- tibble(nonlocal_symbol = rownames(pm[indic_nonlocal, ])[ord_nonlocal], index = 1:length(nonlocal_symbol)) %>%
  inner_join(bar, by = "nonlocal_symbol") %>% 
  left_join(nonlocal_annot, by = c("nonlocal_symbol" = "symbol")) %>%
  ggplot() + geom_point(aes(y = pos, x = index, color = lod_diff > 1.5, nonlocal = nonlocal_symbol, lod_diff = lod_diff))
